Modelling circumstellar discs with 3D radiation hydrodynamics

### 2009-06-29 SPH-Torus-paper1/sphtorus_benchmark.tex
move temperature plots before SEDs in benchmark comparison section.
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@1974 25464f21-6099-402c-b30a-640f1d40d38b

\section{SED generation and temperature distribution}

@@

Further calculations were performed using different numbers of SPH
particles and the enhanced grid resolution described above. SEDs are
plotted in Fig.~\ref{fig:n_part_sed_013} for a viewing angle of 12.5
degrees and Fig.~\ref{fig:n_part_sed_077} for a viewing angle of 77.5
degrees. The density distribution in the central region is plotted in
Fig.~\ref{fig:n_part_central_dens}.
%
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e8}}
  \caption{SEDs at a 12.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_013}
\end{figure*}
%
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e8}}
  \caption{SEDs at a 77.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_077}
\end{figure*}
%
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.22]{figures/dens_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.22]{figures/dens_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.22]{figures/dens_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.22]{figures/dens_1e8}}
  \caption{Central density on AMR grid derived from different numbers
    of SPH particles. The axis units are $10^{13}$~cm.} 
  \label{fig:n_part_central_dens}
\end{figure*}
With $10^5$ and $10^6$ particles a central gap is present but is too
large. With $10^7$ particles there
is more material close in to the source but the gap itself is not
represented and the source is obscured. Only with $10^8$ particles is
a gap of approximately the correct size present.

Figure~\ref{fig:n_part_tem_diff} shows fractional errors in the
temperature distribution for different numbers of particles. {\small{TORUS}}
calculates the temperature distribution using a high resolution 2D
geometry, in order to determine the correct temperature as a function
of cylindrical polar $r$ and $z$ co-ordinates. This is then
compared to the temperature from the SPH disc for each SPH
particle. Positive values of the fractional error indicate that the
SPH disc is hotter than the expected value.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/tdiff_run1}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/tdiff_run2}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/tdiff_run3}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/tdiff_run4}}
  \caption{Fractional temperature error for discs represented by
    different numbers of SPH particles. Axis units are AU.} 
  \label{fig:n_part_tem_diff}
\end{figure*}
When using $10^5$ particles to represent the disc there is a
too much heating in the disc mid plane and insufficient heating
further out of the mid plane. The effect is reduced when $10^6$
particles are used but is only reduced to a low level with $10^8$
particles. The representation of the central region of the disc not
only has a significant impact on the SEDs but also on the temperature
distribution within the disc.

@@

The SPH code uses a non-dimensionalised unit system and the radiative
transfer step is called after an elapsed time of 1.0 code units (1.89
years). This time interval was chosen to allow the hydrodynamics to
respond to the radiative heating and achieve a suitable balance
between the hydrodynamics and the slower radiative transfer
calculation.

>>>>

\section{Temperature distribution and SED generation}

@@

Further calculations were performed using different numbers of SPH
particles and the enhanced grid resolution described
above. Figure~\ref{fig:n_part_tem_diff} shows fractional errors in the
temperature distribution for different numbers of
particles. {\small{TORUS}} calculates the temperature distribution
using a high resolution 2D geometry, in order to determine the correct
temperature as a function of cylindrical polar $r$ and $z$
co-ordinates. This is then compared to the temperature from the SPH
disc for each SPH particle. Positive values of the fractional error
indicate that the SPH disc is hotter than the expected value.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/tdiff_run1}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/tdiff_run2}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/tdiff_run3}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/tdiff_run4}}
  \caption{Fractional temperature error for discs represented by
    different numbers of SPH particles. Axis units are AU.} 
  \label{fig:n_part_tem_diff}
\end{figure*}
When using $10^5$ particles to represent the disc there is a too much
heating in the disc mid plane and insufficient heating further out of
the mid plane. The effect is reduced when $10^6$ particles are used
but is only reduced to a low level with $10^8$ particles.

To allow a more quantitative comparison the mid-plane temperature and
the benchmark mid-plane temperature are plotted in
Fig.~\ref{fig:n_part_tem_diff_midpane} for all particles within 0.01
scale heights of the mid-plane. The fractional temperature error for
each particle is also plotted where a positive error indicates a
temperature hotter than the benchmark value.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/mp_comp1}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/mp_comp2}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/mp_comp3}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/mp_comp4}}
  \caption{Fractional temperature error in the mid-plane for discs represented by
    different numbers of SPH particles.} 
  \label{fig:n_part_tem_diff_midpane}
\end{figure*}
SEDs are plotted in Fig.~\ref{fig:n_part_sed_013} for a viewing angle
of 12.5 degrees and Fig.~\ref{fig:n_part_sed_077} for a viewing angle
of 77.5 degrees.

The density distribution in the central region is
plotted in Fig.~\ref{fig:n_part_central_dens}.
%
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e8}}
  \caption{SEDs at a 12.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_013}
\end{figure*}
%
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e8}}
  \caption{SEDs at a 77.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_077}
\end{figure*}
%
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.22]{figures/dens_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.22]{figures/dens_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.22]{figures/dens_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.22]{figures/dens_1e8}}
  \caption{Central density on AMR grid derived from different numbers
    of SPH particles. The axis units are $10^{13}$~cm.} 
  \label{fig:n_part_central_dens}
\end{figure*}
With $10^5$ and $10^6$ particles a central gap is present but is too
large. With $10^7$ particles there is more material close in to the
source but the gap itself is not represented and the source is
obscured. Only with $10^8$ particles is a gap of approximately the
correct size present.  The representation of the central region of the
disc is significantly affected by the number of particles used to
represent the disc. The representation of the central region has a
significant impact on both the temperature distribution within the
disc and the emergent SEDs.

@@

The SPH code uses a non-dimensionalised unit system and the radiative
transfer step is called after an elapsed time of 1.0 code units (1.89
years) which corresponds to 3 hydrodynamics time steps. This time
interval was chosen to allow the hydrodynamics to respond to the
radiative heating and achieve a suitable balance between the
hydrodynamics and the slower radiative transfer calculation.
